---
title: "Sandhills lake color Thiel-Sen trends"
author: "Daniel Gschwentner"
date: "2024-12-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "C:/Users/DanielGschwentner/OneDrive - University of Nebraska-Lincoln/Sandhills LimnoSat lake color v2")
```

```{r}
rm(list = ls())
lapply(c("tidyverse", "trend", "brms", "tidybayes"), require, character.only = T)
source("code/theme_R.r")
```
## Load data

```{r}
# data
limnosat.annual <- read_csv("data/Sandhills_limnosat_annual_medians_17Feb25.csv")
```

## Thiel Sen slope for median DWL

```{r}
# Thiel Sen-slopes for DWL

# individual lakes
out.grid <- expand.grid(Hylak_id = unique(limnosat.annual$Hylak_id),
                        z = NA, p = NA, Sen = NA, sign = NA, dir = NA)
# loop over lakes
for(i in 1:nrow(out.grid)){ # lapply does not modify the original grid so use for loop instead
  # filter data set
  x <- out.grid$Hylak_id[i]
  x <- limnosat.annual[limnosat.annual$Hylak_id == x, ]
  ts.in <- ts(x$dWL, start = 1984, end = 2019)
  sen <- sens.slope(ts.in)
  # save stats
  out.grid[i, "z"] <- sen$statistic[["z"]]
  out.grid[i, "Sen"] <- sen$estimates[["Sen's slope"]]
  out.grid[i, "p"] <- sen$p.value
  out.grid[i, "sign"] <- ifelse(sen$p.value <0.05, "sign.", "ns.") # stat. sign.
  out.grid[i, "dir"] <- ifelse(sen$statistic[["z"]] < 0, "Decr.", 
                               ifelse(sen$statistic[["z"]] == 0, "No change", "Incr."))
}

# tabulate trends
table(out.grid$sign) # significant trend in 166 lakes; 34% of lakes
table(out.grid$sign, out.grid$dir) # 100 lakes are moving towards blue wavelengths while 65 towards browner wavelengths...

# save output file
write_csv(out.grid, "data/outputs/ThielSen_slopes_dwl.csv")

```


## Fit bayesian models to determine probability of a trend

```{r}
# add T/F for stats
# all lakes
out.grid$trend <- ifelse(out.grid$sign == "sign.", 1, 0)
# blueing
out.grid$decreasing <- ifelse(out.grid$dir == "Decr." & out.grid$sign == "sign.", 1, 0)
# greening
out.grid$increasing <- ifelse(out.grid$dir == "Incr." & out.grid$sign == "sign.", 1, 0)

```


```{r}
# fit bayesian model to data sets
# see this guide for bernoulli distribution: https://bayesball.github.io/BRMS/binomial-modeling.html

# set seed
set.seed(123)

# all trends
change.brm <- brm(
  bf(trend~0 + Intercept), 
  data = out.grid,
  family = bernoulli(),
  prior = c(prior("uniform(-1000, 1000)", lb = -1000, ub = 1000)),
  file  = "model_files/prob_trend_dwl",
  chains = 4, iter = 8000, warmup = 1000
)
plot(change.brm)
pp_check(change.brm)
summary(change.brm)
# calculate posterior probability
prob.change <- posterior_samples(change.brm)
prob.change <- prob.change %>% mutate(p = exp(b_Intercept)/(1 + exp(b_Intercept)))
quantile(prob.change$p, c(0.025, 0.5, 0.975)); max(prob.change$p); min(prob.change$p)
hist(prob.change$p)

# greens
green.brm <- brm(
  bf(decreasing~0 + Intercept), 
  data = out.grid,
  family = bernoulli(),
  prior = c(prior("uniform(-1000, 1000)", lb = -1000, ub = 1000)),
  file  = "model_files/prob_trend_dwl_green",
  chains = 4, iter = 8000, warmup = 1000 
)
plot(green.brm)
pp_check(green.brm)
summary(green.brm)
# calculate posterior probability
prob.green <- posterior_samples(green.brm)
prob.green <- prob.green %>% mutate(p = exp(b_Intercept)/(1 + exp(b_Intercept)))
quantile(prob.green$p, c(0.025, 0.5, 0.975))
hist(prob.green$p)

# browns
browns.brm <- brm(
  bf(increasing~0 + Intercept), 
  data = out.grid,
  family = bernoulli(),
  prior = c(prior("uniform(-1000, 1000)", lb = -1000, ub = 1000)),
  file  = "model_files/prob_trend_dwl_brown",
  chains = 4, iter = 8000, warmup = 1000 
)
plot(browns.brm)
pp_check(browns.brm)
summary(browns.brm)
# calculate posterior probability
prob.browns <- posterior_samples(browns.brm)
prob.browns <- prob.browns %>% mutate(p = exp(b_Intercept)/(1 + exp(b_Intercept)))
quantile(prob.browns$p, c(0.025, 0.5, 0.975))
hist(prob.browns$p)

# combine draws for posterior probability
prob.trend <- bind_rows(prob.green, prob.browns)
prob.trend$dir <- c(rep("Decr.", nrow(prob.green)), rep("Incr.", nrow(prob.browns)))
# data frame of stats for figure
prob.stats <- data.frame(
  value = c(quantile(prob.green$p, c(0.025, 0.5, 0.975)), 
            quantile(prob.browns$p, c(0.025, 0.5, 0.975))),
  dir = rep(c("Decr.", "Incr."), each = 3), 
  cat = rep(c("lwr", "median", "upr"), 2)
)

# how much more probable is greening vs. browning?
quantile(prob.green$p, c(0.025, 0.5, 0.975))/quantile(prob.browns$p, c(0.025, 0.5, 0.975))

```
## Figure of model output/summary

```{r}
# figure of trends; only showing trends for data including lake color
trends.data <- merge(limnosat.annual, out.grid, by = "Hylak_id")

# summary trends data
trends.data %>% filter(trend == 1) %>%  group_by(dir) %>% summarise(median = median(Sen),
                                                                    min = min(Sen), max = max(Sen),
                                                                    q025 = quantile(Sen, 0.025),
                                                                    q975 = quantile(Sen, 0.975))

(fig.trends <- trends.data %>%
  filter(sign == "sign.") %>%
  ggplot() + 
  geom_line(aes(year,dWL, group = Hylak_id, col = dir), alpha = 0.3) + 
  #facet_wrap(dir~., nrow = 2, ncol = 1) + 
  scale_color_manual(values = c("Decr." = "forestgreen", "Incr." = "brown")) + 
  #ggsci::scale_color_jco()  +
  labs(x = "Year", y = "DWL") + 
  scale_x_continuous(breaks = c(1990, 2000, 2010, 2019)) + 
  theme(legend.position = "none"))

(fig.slopes <- trends.data %>%
  filter(sign == "sign.") %>%
  ggplot() + 
  geom_density(aes(Sen, col = dir)) + 
  geom_vline(aes(xintercept = 0), lty = "dashed") + 
  scale_color_manual(values = c("Decr." = "forestgreen", "Incr." = "brown")) + 
  #ggsci::scale_color_jco()  +
  #facet_wrap(dir~., nrow = 2, ncol = 1) + 
  labs(x = expression("Sen's slope nm year"^-1), y = "Density", col = NULL) + 
  theme(legend.position = c(0.2, 0.6)))

(fig.probs <- prob.trend %>%
  ggplot() + 
  geom_density(aes(p * 100, col = dir)) + 
  geom_vline(inherit.aes = F, data = prob.stats, 
             aes(xintercept = value * 100, col = dir), lty = "dashed") + 
   scale_color_manual(values = c("Decr." = "forestgreen", "Incr." = "brown")) + 
  #ggsci::scale_color_jco()  +
  #facet_wrap(dir~., nrow = 2, ncol = 1) + 
  theme(legend.position = "none") + 
  labs(x = "Probability (%)", y = "Density"))

fig.trends.slopes <- ggpubr::ggarrange(
  plotlist = list(fig.trends, fig.slopes, fig.probs), ncol = 1, nrow = 3, 
  labels = letters[1:3], align = "hv") #, heights = c(0.3, 0.2, 0.2))
fig.trends.slopes

ggsave("figures/dwl_thiel_sen_trends.jpg", fig.trends.slopes, width = 3.5, height = 6, dpi = 400)

```

## Lake color over time plot

I was curious how lake color for individual lakes was changing over time and decided to make a large timeseries plot of lake color groups; did this for annual median lake color and monthly values too.


Annual median lake color

```{r}
annual.col <- limnosat.annual[,c("Hylak_id", "year", "dWL")] %>%
  mutate(color.group = ifelse(dWL <= 495, "Blue", ifelse(dWL >= 560, "Brown", "Green"))) %>%
  group_by(Hylak_id) %>% mutate(mean.dwl = mean(dWL)) %>% arrange(mean.dwl) %>% # arrange from lowest to highest mean DWL
  mutate(Hylak_id = as.factor(Hylak_id), Hylak_id = fct_reorder(Hylak_id, mean.dwl))
# plot color classes
(annual.chart <- annual.col %>% ggplot(aes(year, Hylak_id, fill = color.group)) + 
  geom_raster() + scale_fill_manual(values =c ("Blue" = "blue", "Brown" = "brown", "Green" = "forestgreen")) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank()) + 
  scale_x_continuous() + labs(x = NULL, y = NULL, fill = NULL) + scale_y_discrete(limits = rev))

```

Create same figure for monthly values but only restrict to the lakes for which annual data is available

```{r}
# load data set and filter
monthly.col <- read_csv("data/Sandhills_limnosat_no_winter_17Feb25.csv") %>%
  filter(Hylak_id %in% annual.col$Hylak_id) %>%
  mutate(Hylak_id = factor(Hylak_id, levels = levels(annual.col$Hylak_id)), month2 = month(date, label = T)) %>%
  select(Hylak_id, year, month2, dWL) %>%
  mutate(color.group = ifelse(dWL <= 495, "Blue", ifelse(dWL >= 560, "Brown", "Green")))

(monthly.chart <- monthly.col %>% 
  ggplot(aes(year, Hylak_id, fill = color.group)) + 
  geom_raster() + scale_fill_manual(values =c ("Blue" = "blue", "Brown" = "brown", "Green" = "forestgreen")) + 
  facet_wrap(month2~.) + theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line.y = element_blank(), 
                               panel.margin.x=unit(0.0, "lines") , panel.margin.y=unit(0.0,"lines")) + 
  scale_x_continuous() + labs(x = NULL, y = NULL, fill = NULL) + scale_y_discrete(limits = rev))

```
Combine chart

```{r}
# patch together
chart <- ggpubr::ggarrange(
  plotlist = list(annual.chart, monthly.chart), widths = c(3/7, 4/7), align = "hv", ncol = 2, labels = letters[1:2], common.legend = T, legend = "bottom"
)
# save
chart
ggsave("figures/color_timeseries_chart.jpg", chart, width = 10, height = 5, dpi = 500)
```

