---
title: "Merge Groundwater level and DWL data"
author: "Daniel Gschwentner"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(root.dir = "C:/Users/DanielGschwentner/OneDrive - University of Nebraska-Lincoln/Sandhills LimnoSat lake color v2")
```

```{r}
rm(list = ls())
library(tidyverse)
library(sf)
source("code/theme_R.r")
```

## Load data sets

Limnosat DWL data

```{r}
# data
sh.annual <- read_csv("data/Sandhills_limnosat_annual_medians_17Feb25.csv")
sh.monthly <- read_csv("data/Sandhills_limnosat_no_winter_17Feb25.csv")
``` 
## Geospatial analysis

Note to myself on 18-Feb-25: I decided to make the data wrangling more reproduceable as there were some issues with disappearing lakes when I reassembled the limnosat data set. The code now selects lakes that are within the 20 km buffer of the selected groundwater wells AND includes the lakes from VLNWR (these are manually added).

Load GIS data

```{r}
# wells
wells.2.keep <- st_read(dsn = "data/GIS/sandhill_lakes.gpkg", layer = "final_groundwater_wells")  # file with 9 wells for which we have GW data
# wells buffer
wells.2.keep.buffer <- st_read(dsn = "data/GIS/sandhill_lakes.gpkg", layer = "gw_wells_20km_buffer") %>% # gotta filter this because there are one or two extra buffers for some reason
  filter(SiteNo %in% wells.2.keep$SiteNo)
# valentine National Valetine Wildlife Refuge
vlnwr <- st_read(dsn = "data/GIS/sandhill_lakes.gpkg", layer = "usgs_refuges" ) %>%
  filter(ORGNAME == "VALENTINE NATIONAL WILDLIFE REFUGE")
# all lakes
lakes <- read_csv("data/Sandhills_limnosat_annual_medians_17Feb25.csv") %>%
  dplyr::select(Hylak_id, lon, lat) %>% distinct() %>%
  st_as_sf(., coords = c("lon", "lat"))
st_crs(lakes) <- 3857
```

```{r}
# check overlap of data <-- this looks good so we can proceed with lake selection
ggplot() + 
  geom_sf(data = wells.2.keep.buffer) +
  geom_sf(data = vlnwr) + 
  geom_sf(data = lakes)
```
## GIS wrangling

Select lakes in buffers and VLNWR
```{r}
# lakes in the buffer
buffer.lakes <- st_filter(lakes, wells.2.keep.buffer, .predicate = st_within)
buffer.lakes <- st_join(buffer.lakes, wells.2.keep.buffer, .predicate = st_within)

# lakes in vlnwr
vlnwr.lakes <- st_filter(lakes, vlnwr, .predicate = st_within)
vlnwr.lakes <- vlnwr.lakes %>% mutate(SiteNo = 40, Name =  "SECherryCounty", lon = -100.5313, lat = 42.21729)

# merge
selected.lakes <- rbind(buffer.lakes, vlnwr.lakes)

```

Distance matrix between lakes and wells

```{r}

## Generate distance matrix

# distance matrix is in units m
d.matrix <- st_distance(selected.lakes, wells.2.keep) %>% as.data.frame()
# add Hylak_id
d.matrix$Hylak_id <- selected.lakes$Hylak_id
# wide to long and some formatting
d.matrix <- d.matrix %>%
  gather("SiteNo", "dist_m", -Hylak_id) %>%
  mutate(SiteNo = rep(c(22, 27, 44, 39, 46, 36, 8, 40, 13), each = nrow(selected.lakes)),
         dist_km = units::drop_units(dist_m)/1000)
```

Merge data sets
```{r}
## Add distance matrix to selected lakes
selected.lakes <- merge(selected.lakes, d.matrix, by = c("Hylak_id", "SiteNo"))

```

Overview of the selected lakes per well and mean distances
```{r}
selected.lakes.summary <- selected.lakes %>% group_by(Name) %>%
  summarise(n = n(), mean_dist_km2 = mean(dist_km), min_dist_km2 = min(dist_km), max_dist_km2 = max(dist_km))
mean(selected.lakes.summary$n); mean(selected.lakes.summary$mean_dist_km2)
```

Save output

```{r}
#st_write(selected.lakes, dsn = "data/GIS/sandhill_lakes.gpkg", layer = "selected_gw_lakes_18Feb25")
selected.lakes <- sf::st_read(dsn = "data/GIS/sandhill_lakes.gpkg", layer = "selected_gw_lakes_18Feb25")
```
## Load groundwater data
Load groundwater data (previously cleaned)

```{r}
# load groundwater data
gw.input <- read_csv("data/GWL/sandhills_gwl_monthly_clean.csv")
# filter
gw.input <- gw.input %>% filter(SiteNo %in% selected.lakes$SiteNo)
# aggregate on an annual base
gw.annual <- gw.input %>% 
  mutate(year = year(date)) %>% 
  group_by(SiteNo, Name, year) %>%
  summarise(gwl_bl_surface_ft = mean(gwl_bl_surface_ft))
#write_csv(gw.annual, "data/annual_gw_levels.csv")
```

```{r}
# thiel-sen slopes for GW
# sheridan
sheridan <- gw.annual[gw.annual$Name == "SheridanSouth",]
sheridan <- ts(sheridan$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
sheridan.sen <- trend::sens.slope(sheridan)
sheridan.sen

# antioch
antioch <- gw.annual[gw.annual$Name == "Antioch",]
antioch <- ts(antioch$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
antioch.sen <- trend::sens.slope(antioch)
antioch.sen

# Crescent Lake
crescent <- gw.annual[gw.annual$Name == "CrescentLake",]
crescent <- ts(crescent$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
crescent.sen <- trend::sens.slope(crescent)
crescent.sen

# SW Cherry County Lake
swcherry <- gw.annual[gw.annual$Name == "SWCherryCounty",]
swcherry <- ts(swcherry$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
swcherry.sen <- trend::sens.slope(swcherry)
swcherry.sen

# Grant County
grant <- gw.annual[gw.annual$Name == "GrantCounty",]
grant <- ts(grant$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
grant.sen <- trend::sens.slope(grant)
grant.sen

# Cherry County
cherry <- gw.annual[gw.annual$Name == "CherryCounty",]
cherry <- ts(cherry$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
cherry.sen <- trend::sens.slope(cherry)
cherry.sen

# Arthur County
arthur <- gw.annual[gw.annual$Name == "ArthurCounty",]
arthur <- ts(arthur$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
arthur.sen <- trend::sens.slope(arthur)
arthur.sen

# SE Cherry County
SECherry <- gw.annual[gw.annual$Name == "SECherryCounty",]
SECherry <- ts(SECherry$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
SECherry.sen <- trend::sens.slope(SECherry)
SECherry.sen

# Barta Brothers
barta <- gw.annual[gw.annual$Name == "BartaBrothers",]
barta <- ts(barta$gwl_bl_surface_ft[1:12], start = 2008, end = 2019) 
barta.sen <- trend::sens.slope(barta)
barta.sen

```


```{r}
# arrange groundwater levels from W to E
gw.annual$Name <- factor(gw.annual$Name, levels = c("SheridanSouth", "Antioch", "CrescentLake", "SWCherryCounty", "GrantCounty", "CherryCounty", "ArthurCounty", "SECherryCounty", "BartaBrothers"))

# add in sign. from Thiel-Sen
gw.annual$sign <- ifelse(gw.annual$Name %in% c("CherryCounty", "SECherryCounty", "BartaBrothers"), "n.s.", "sign.")

scaleFUN <- function(x) sprintf("%.0f", x)
# plot groundwater trends
(gw.raw <- gw.annual %>%
  ggplot(aes(year, gwl_bl_surface_ft, group = SiteNo, col = Name, alpha = sign)) + 
  geom_hline(yintercept = 0, lty = "dashed", col = "black") + 
  geom_line(lwd = 0.75) + geom_point(pch = 21, fill = "white", size = 2, stroke = 0.75) + 
  labs(x = NULL, y = "Groundwater level \n ft. below surface", col = NULL) + 
  scale_x_continuous(limits = c(2008, 2019), labels = scaleFUN) + scale_y_reverse() + 
  ggsci::scale_color_jco() + theme(legend.position = "right") )
ggsave("figures/gw_trends.jpg", gw.raw, width = 7, height = 3.5, dpi = 500)

```

```{r}
# descriptive stats
gw.annual %>% filter(year < 2019) %>% 
  group_by(Name) %>% 
  summarise(mean = mean(gwl_bl_surface_ft),
            min = min(gwl_bl_surface_ft),
            max = max(gwl_bl_surface_ft)) %>% 
  arrange(mean)
```

## Data aggregation and filtering

```{r}
# select groundwater lakes
limnosat.gw <-  sh.annual %>% filter(Hylak_id %in% selected.lakes$Hylak_id)
# add gwl-ID
limnosat.gw <- merge(limnosat.gw, selected.lakes[,c("Hylak_id", "SiteNo", "Name", "dist_km")], by = "Hylak_id")
```


```{r}
# merge data sets
dwl.gwl <- merge(limnosat.gw, gw.annual, by = c("year","Name", "SiteNo"), all.x = F, all.y = F)

```
Save output

```{r}
#write_csv(dwl.gwl, "data/Sandhills_limnosat_annual_medians_gw_18Feb25.csv")
```


## Aggregate on a monthly basis
(for fun and testing)

```{r}
# aggregate GW on monthly basis
gw.monthly <- gw.input %>%
  mutate(year = year(date), month = month(date), date = ymd(paste0(year, "/", month, "/01"))) %>%
  group_by(SiteNo, Name, date) %>% 
  summarise(gwl_bl_surface_ft = mean(gwl_bl_surface_ft))

```



```{r}
# arrange groundwater levels from W to E
gw.monthly$Name <- factor(gw.monthly$Name, levels = c("SheridanSouth", "Antioch", "CrescentLake", "SWCherryCounty", "GrantCounty", "CherryCounty", "ArthurCounty", "SECherryCounty", "BartaBrothers"))

scaleFUN <- function(x) sprintf("%.0f", x)
# plot groundwater trends
(gw.raw2 <- gw.monthly %>%
  ggplot(aes(date, gwl_bl_surface_ft, group = SiteNo, col = Name)) + 
  geom_hline(yintercept = 0, lty = "dashed", col = "black") + 
  geom_line(lwd = 0.75) +  #geom_point(pch = 21, fill = "white", size = 2, stroke = 0.75) + 
  labs(x = NULL, y = "Groundwater level \n ft. below surface", col = NULL) + 
  scale_x_continuous(limits = as.Date(c("2008-01-01", "2020-01-01")), labels = scaleFUN) + 
  scale_y_reverse() + 
  ggsci::scale_color_jco() + theme(legend.position = "right") )
#ggsave("figures/gw_trends.jpg", gw.raw, width = 7, height = 3, dpi = 500)

```


```{r}
# select groundwater lakes
limnosat.gw2 <-  sh.monthly %>% filter(Hylak_id %in% selected.lakes$Hylak_id)
# add gwl-ID
limnosat.gw2 <- merge(limnosat.gw2, selected.lakes[,c("Hylak_id", "SiteNo", "Name", "dist_km")], by = "Hylak_id")
```


```{r}
# merge data sets
dwl.gwl2 <- merge(limnosat.gw2, gw.input, by = c("date","Name", "SiteNo"), all.x = F, all.y = F)

```
Save output

```{r}
#write_csv(dwl.gwl2, "data/Sandhills_limnosat_monthly_gw_18Feb25.csv")
```



